### 2022-03-11 bools/bools.tex
intro

Our approach is implemented in the Zipperposition theorem prover
\cite{sc-15-simon-phd,sc-supind-17}. This prover was used by Bentkamp et al.\
to implement \lsup{}. We further extend
their implementation.

@@

As mentioned in Chapter~\ref{ch:unif}, Zipperposition is an open source prover
written in OCaml. From its inception, it was designed as a prover that supports
easy extension of its base superposition calculus to various theories, including
arithmetic, induction and limited support for higher-order logic
\cite{sc-15-simon-phd}.

In Zipperposition, applications are represented in flattened, spine notation (Sect.~\ref{sec:ehoh:types-and-terms}).
This means that higher-order term $(\cst{f} \, \cst{a}) \, cst{b}$ is
represented the same as first-order term $\cst{f}(\cst{a}, \cst{b})$. In
addition, Zipperposition uses associativity of $\iand$ and $\ior$ to flatten out
the nested applications of these symbols. For example, terms $\cst{p} \iand
(\cst{q} \iand \cst{r})$ and $(\cst{p} \iand \cst{q}) \iand \cst{r}$ are
internally stored as $\iand \, \cst{p} \, \cst{q} \, \cst{r}$. Zipperposition's
support for $\lambda$-terms is used to represent quantified nested Booleans:
Formulas $\iforall x. \, \, f$ and $\iexists x. \, f$ are represented as
$\iforall \, (\lamx{f})$ and $\iexists \, (\lamx{f})$. After clausification of
the input problem, no nested Booleans will be modified or renamed using fresh
predicate symbols.

The version of Zipperposition preceding our modifications distinguished between non\-equational
and equational literals. Following E \cite{scv-19-e23}, we
modified Zipperposition to represent all literals equationally: a nonequational
literal $f$ is stored as $f \eq \itrue$, whereas $\neg f$ is stored as $f
\noteq \itrue$. Equations of the form $f \eq \ifalse $ and $f \noteq \ifalse$ are
transformed into $f \noteq \itrue$ and $f \eq \itrue$, respectively.

@@

\looseness=-1
Rule \infname{PI} was already implemented by Simon Cruanes in Zipperposition,
before we started our modifications. Its implementation contains the following modes that generate
sets of possible terms $f$ for $P: \tau_1 \rightarrow \cdots \rightarrow \tau_n
\rightarrow o$: \emph{Full}, \emph{Pragmatic}, and $\mathit{Imit}_{\star}$. In  $\mathit{Imit}_{\star}$,
$\star$ is an element of a set of logical constants $S = \{ \iand, \ior,
\mathop{\ieq}\typeargs{\alpha}, \inot, \iforall, \iexists \}$. Mode \emph{Full}
generates imitations (for $P$) of all elements of $S$. Mode $\emph{Pragmatic}$
generates imitations of $\inot$, $\itrue$ and $\ifalse$; if there exist indices $i,j$
such that $i\ineq j$ and  $\tau_i = \tau_j$, it generates $\lam{\tuplen{x}}{x_i
\ieq x_j}$; if there exist indices $i,j$ such that $i \ineq j$, and $\tau_i =
\tau_j = o$, then it generates $\lam{\tuplen{x}}{x_i \iand x_j}$ and
$\lam{\tuplen{x}}{x_i \ior x_j}$; if for some $i$, $\tau_i = o$, then it
generates $\lam{\tuplen{x}}{x_i}$. Mode $\mathit{Imit}_\star$ generates
imitations of $\itrue$, $\ifalse$ and $\star$. In addition, $\mathit{Imit_{\iforall\iexists}}$ which generates imitations
of both $\iforall$ and $\iexists$.

@@

\looseness=-1
The axiom of choice is notoriously difficult to handle efficiently in higher-order provers. Andrews formulates
this axiom as $\iforall p.\, (\iexists
x.\, p \, x) \iimplies p \, (\varepsilon \, p)$, where $\varepsilon :
\forallty{\alpha}(\alpha \rightarrow o) \rightarrow \alpha$ denotes the \emph{choice
operator} \cite{pa-01-classical-ty-thy}. After clausification, this axiom becomes $P \, X \noteq \itrue \llor P
\, (\varepsilon \, P) \eq \itrue$. Since the term $P \, X$ matches any Boolean
term in the proof state, this axiom is very explosive. Therefore, Leo-III
\cite{sb-21-leo3} heuristically recognizes symbols that correspond to choice and deals with
them on the calculus level. Namely, whenever a clause $C = P \, X \noteq \itrue \llor P \, (\cst{f} \,
P) \eq \itrue$ is chosen for processing, $C$ is removed from the proof state and $\cst{f}$ is
added to set of choice functions $\mathit{CF}$ (which initially contains just
$\varepsilon$). Later, elements of $\mathit{CF}$ will be used to heuristically
instantiate the axiom of choice. We reused the method of recognizing choice
functions, but generalized the rule for creating the instance of
the axiom (assuming $\xi \in \mathit{CF}$ and $X$ and $z$ are fresh variables):
%
$$\namedinference{Choice}
               {C[\xi \, t]}
               { X \, (t \, Y) \noteq \itrue \llor X \, (t \, (\xi \, (\lam{z}{X \, (t \, z)})) ) \eq \itrue }$$
%
Let $D$ be the conclusion of \infname{Choice}. The fresh variable $X$ in $D$ acts as
arbitrary context around $t$, the chosen instantiation for $P$ from axiom of choice;
the variable $X$ can later be replaced by imitations of logical symbols to create more
complex instantiations of the choice axiom. To generate useful instances early, we create two instances: $\substcl{\{X
\mapsto \lam{z}{z}\}}{D}$ and $\substcl{\{X \mapsto \lam{z}{\inot\, z}\}}{D}$. Then, depending on
Zipperposition options, $D$ will either be deleted or kept. Note that $D$
will not subsume its instances, since the matching algorithm Zipperposition uses is
too weak for this \cite[Sect.~6]{bbtvw-21-sup-lam}
\looseness=-1

Most provers natively support both functional and Boolean extensionality reasoning: Bhayat et al. \cite{br-20-full-sup-w-combs} modify
first-order unification to return unification constraints consisting of pairs of
terms of functional type, whereas Steen relies on the unification rules of
Leo-III's calculus \cite[Section 4.3.3]{as-18-phd} to deal with extensionality.
As a pragmatic extension of the \lsup{} calculus, Bentkamp et al. \cite{bbtvw-21-sup-lam} propose to alter core generating inference rules of \lsup{}
to support functional extensionality. Instead of requiring that terms
involved in the inference are unifiable, it is required that they can be
decomposed into \emph{disagreement pairs} such that at least one of the
disagreement pairs is of functional type. Disagreement pairs of terms $s$ and
$t$ of the same type are defined inductively using function $\textsf{dp}$:
$\textsf{dp}(s,t) = \emptyset$ if $s$ and $t$ are equal; $\textsf{dp}(a \,
\tuplen{s}, b \, \tuple{t}{m}) = \{(a \,
\tuplen{s}, b \, \tuple{t}{m})\}$ if $a$ and $b$ are different heads;
$\textsf{dp}(\lam{x}{s}, \lam{y}{t}) = \{ (\lam{x}{s}, \lam{y}{t}) \}$;
$\textsf{dp}(a \, \tuplen{s}, a \, \tuplen{t}) = \bigcup_{i=1}^{n}
\textsf{dp}(s_i, t_i)$. Then the extensionality rules are stated as follows:
\pagebreak[2]
\begin{align*}
  & \namedinference{AbsSup}
  {s \eq t \llor C \qquad u[s'] \mathrel{\dot{\eq}} v \llor D}
  {\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \llor  u[t] \mathrel{\dot{\eq}} v \llor C \llor D}} \\
  &\namedinference{AbsER}{s \noteq s' \llor C}{\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \cdots \llor  C}} \\
  &\namedinference{AbsEF}{s \eq t \llor  s' \eq u \llor C}{\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \llor t \noteq u \llor  s' \eq u \llor C}}
\end{align*}

\noindent\looseness=-1
In each of the rules, $\sigma$ is a most general unifier of the types of $s$ and
$s'$, and $\textsf{dp}(\sigmaterm{s} ,\sigmaterm{s'}) = \{ (s_1, s'_1), \ldots,
(s_n, s'_n)\}$. All side conditions for extensional rules are the same as for
the standard rules, except that condition that $s$ and $s'$ are unifiable is
replaced by the condition that at least one $s_i$ is of functional type and that
$n>0$. Rules \infname{AbsSup}, \infname{AbsER}, and \infname{AbsEF} are
extensional versions of superposition, equality resolution and equality
factoring (Sect.~\ref{sec:pre:rules}). By \infname{Abs} we denote the union of
these three rules. We extend this rule to support Boolean extensionality by
requiring that at least one $s_i$ is of functional or type $o$, and adding the
condition ``$\textsf{dp}(f, g) = \{(f,g)\}$ if $f$ and $g$ are different
formulas'' to the definition of $\textsf{dp}$. If another condition of \textsf{dp}'s 
definition is also applicable, the newly added one is preferred.

Consider the clause $\cst{f} \, (\inot\cst{p} \ior \inot\cst{q}) \noteq \cst{f} \,
(\inot (\cst{p} \iand \cst{q}))$. This problem is obviously unsatisfiable, since
the arguments of $\cst{f}$ on the different sides of the disequation are Boolean-extensionally
equal. Without the \infname{Abs} rules Zipperposition will rely on
\infname{Cases}(\infname{Simp}) and \infname{IC} rules to derive the empty
clause. On the other hand, \infname{AbsER} will generate~$C = \inot\cst{p} \ior \inot\cst{q} \noteq
\inot (\cst{p} \iand \cst{q})$. Then, $C$ will get clausified using
\infname{IC}, effectively reducing the problem to $\inot (\inot\cst{p} \ior
\inot\cst{q} \iequiv \inot (\cst{p} \iand \cst{q}))$, which is first-order.

@@

The winner of the higher-order theorem division of the 2019 edition of CASC \cite{gs-19-casc27}, Satallax \cite{cb-12-satallax}, has one
crucial advantage over Zipperposition: it is based on higher-order
tableaux, and as such it does not require formulas to be converted to clauses. The advantage of tableaux is that once it instantiates a variable with a
term, this instantiation naturally propagates through the whole formula. In
Zipperposition, which is based on \lsup{}, the original formula
is clausified and instantiating a variable in a clause $C$ does not automatically instantiate
it in all clauses that are results of clausification of the same formula as $C$.
To mitigate this issue, we have created extensions of equality resolution and equality factoring
that take Boolean extensionality into account:
%
\begin{align*}
  & \namedinference{BoolER}
  { s \eq s' \llor C}
  {\sigmacl{C}}
  &&
  \namedinference{BoolEF$+-$}
  {P\,\tuplen{s} \eq \itrue \llor s' \not\eq \itrue \llor C}
  {\sigmacl{s' \not\eq \itrue \llor C}} \\
  & \namedinference{BoolEF$-+$}
  {P\,\tuplen{s} \not\eq \itrue \llor s' \eq \itrue \llor C}
  {\sigmacl{s' \eq \itrue \llor C}} &&
  \namedinference{BoolEF$-$$-$}
  {P\,\tuplen{s} \not\eq \itrue \llor s' \not\eq \itrue \llor C}
  {\sigmacl{s' \not\eq \itrue \llor C}}
\end{align*}
%
All side conditions except for the ones concerning the unifiability of terms are
as in the original equality resolution and equality factoring rules. In rule
\infname{BoolER}, $\sigma$ is a unifier of $ s $ and $ \inot s' $. In the $+-$
and $-+$ versions of $\infname{BoolEF}$, $\sigma$ unifies $ P \, \tuplen{s}
$ and $\inot s'$, and in the remaining version it unifies $ P \, \tuplen{s} $ and
$s'$. Intuitively, these rules bring Boolean (dis)equations in the appropriate
form for application of the corresponding base rules. In particular,
for \infname{BoolER} we interpret the equation $s \eq s'$ as $s \not\eq \inot s'$ which
allows us to simulate \infname{ER} inference from Sect.~\ref{sec:pre:rules}. 
For the \infname{BoolEF} family of rules, we convert disequations into equations
using the trick of negating one side to simulate \infname{EF}.
% It suffices to consider 
% literals of the form $s \eq s'$ for \infname{BoolER} as Zipperposition rewrites
% $s \iequiv t \eq \itrue$ and $\inot(s \iequiv t) \not\eq \itrue$ to $s \eq t$ (and does analogous
% rewriting into $ s \not\eq t $).
The rule \infname{BoolER} only considers literals of the form $s \eq t$. 
Equivalences $s \iequiv t = \itrue$ and disequivalences $(s \iequiv t) \not\eq \itrue$ are automatically simplified to $s \eq t$ or $s \not\eq t$.
The example  \verb|SET557^1| in Sect.~\ref{sect:bool:examples} illustrates how 
\infname{Bool} family of rules helps solve the problems that get obfuscated by clausification.

@@

\looseness=-1
In \infname{DC}$_\eq$ we require both $f$ and $g$ to be formulas and at least one of
them not to be $\itrue$. In \infname{DC}$_{\iforall}$, $Y$ is a fresh variable, and in
\infname{DC}$_{\iexists}$, $\cst{sk}$ is a fresh symbol and $\tuple{\alpha}{}$ and $\tuplen{Y}$ are all the type and term variables occurring
freely in $\iexists x.\, f$. The rules described above are as given by Ganzinger and Stuber (adapted to
our setting), with the omission of rules for negative literals (of the form $ f \not\eq
\itrue$), which are easy to derive and which can be found in their work \cite{gs-05-boolsup}.

\looseness=-1
Naive application of the \infname{DC} rules can result in exponential blowup in
problem size. To avoid this, we rename formulas $f$ that have repeated
occurrences by introducing predicates $\cst{p} \, \tuple{X}{n}$ for them, where
$\tuplen{X}$ are all free variables of $f$. We keep the count of all nonatomic
formulas occurring as either side of a literal. Before applying the \infname{DC}
rules on a clause $f \mathrel{\dot{\eq}} \itrue \llor C $, we check whether the
number of occurrences of $f$  exceeds the threshold $k$. If it does, based on the
polarity of the literal $f \mathrel{\dot{\eq}} \itrue$, we add the clause
$\cst{p} \, \tuplen{Y} \not\eq \itrue \llor f \eq \itrue$ (if the literal is
positive) or $\cst{p} \, \tuplen{Y} \eq \itrue \llor f \not\eq \itrue$ (if the
literal is negative), where $\tuplen{Y}$ are all free variables of $f$ and
$\cst{p}$ is a fresh symbol. Then, we replace the clause $f \mathrel{\dot\eq}
\itrue \llor C$ by $ \cst{p} \, \tuplen{Y} \mathrel{\dot\eq} \itrue \llor C$.

Before the number of occurrences of $f$ is checked, we first check (using a
fast, incomplete matching algorithm, used for simplification and subsumption) if there is a
formula $g$, for which definition was already introduced, such that
$\sigmaterm{g} = f$, for some substitution $\sigma$. This check can have three
outcomes. First, if the definition $\textsf{q} \, \overline{X}_n$ is already
introduced for $g$ with the polarity matching that of $f \, \dot{\eq} \,
\itrue$, then $f$ is replaced by $\sigmaterm{\cst{q} \, \tuplen{X}}$. Second, if
the definition was introduced, but with different polarity, we create the clause
defining $g$ with the missing polarity, and replace $f$ with $\sigmaterm{\cst{q}
\, \tuplen{X}}$. Last, if the there is no renamed formula $g$ generalizing $f$,
then we check if the number of occurrences of $f$ exceeds the threshold.

@@

\looseness=-1
Rules \infname{Cases} and \infname{CasesSimp} deal with Boolean terms,
but we need to rely on extensionality reasoning to deal with $\lambda$-abstractions
whose body has type $o$. Using the observation that the formula $\iforall \tuplen{x}. \, f$ 
implies that $\lam{\tuplen{x}}{f}$ is functional-extensionally equal to $\lam{\tuplen{x}}{\itrue}$ (and similarly, if
$\iforall \tuplen{x}. \, \inot f$, then $\lam{\tuplen{x}}{f} \eq \lam{\tuplen{x}}{\ifalse})$, we
designed the following rule (where $\tuplen{x}$ consists of all loosely bound variables of $f$ ):
%\pagebreak[2]
%
$$ \namedinference{Interpret$\lambda$}{C[\lam{\tuplen{x}}{f}]}
{(\iforall {\tuplen{x}}.\, f) \noteq \itrue \llor C[\lam{\tuplen{x}}{\itrue}] \qquad
(\iforall {\tuplen{x}}.\, \inot f) \noteq \itrue \llor C[\lam{\tuplen{x}}{\ifalse}]} $$
%

@@

The TPTP library contains thousands of higher-order benchmarks, many of them hand-crafted
to point out subtle interferences of functional and Boolean properties of higher-order logic. In this section
we discuss some problems from the TPTP library \cite{gs-17-tptp} that illustrate the advantages and disadvantages
of our approach.

@@

\looseness=-1
Combining these two rules with dynamic clausification is more useful: When rule
\infname{IC} is replaced by rule \infname{DC}, the success rate increases
to 1660 problems, compared to 1646 problems solved by $b$. We also discovered
that reasoning with choice is useful: When rule \infname{Choice} is enabled, the
success rate increases to 1653. We determined that including or excluding the
conclusion $D$ of \infname{Choice}, after it is simplified, makes no difference.
Counterintuitively, disabling \infname{BoolSimp} rule results in 1640 problems,
which is only 6 problems short of configuration $b$. Disabling \infname{Abs} and
\infname{Interpret$\lambda$} rules results in solving 25 and 31 problems less,
respectively. The raw data show that in total, using configurations from Figure
\ref{fig:cases} to Figure \ref{fig:bool-rules}, 1682 problems can be solved.

>>>>

Our approach is implemented in Zipperposition
\cite{sc-15-simon-phd,sc-supind-17}. This theorem prover was used by Bentkamp et al.\
to implement \lsup{}. We further extend
their implementation.

@@

As mentioned in Chapter~\ref{ch:unif}, Zipperposition is an open source prover
written in OCaml. From its inception, it was designed as a prover that supports
easy extension of its base superposition calculus to various theories, including
arithmetic, induction and limited support for higher-order logic
\cite{sc-15-simon-phd,sc-supind-17}.

In Zipperposition, applications are represented in flattened, spine notation (Sect.~\ref{sec:ehoh:types-and-terms}).
This means that the higher-order term $(\cst{f} \, \cst{a}) \, \cst{b}$ is
represented the same as the first-order term $\cst{f}(\cst{a}, \cst{b})$. In
addition, Zipperposition uses associativity of $\iand$ and $\ior$ to flatten out
the nested applications of these symbols. For example, terms $\cst{p} \iand
(\cst{q} \iand \cst{r})$ and $(\cst{p} \iand \cst{q}) \iand \cst{r}$ are
internally stored as $\iand \, \cst{p} \, \cst{q} \, \cst{r}$. Zipperposition's
support for $\lambda$-terms is used to represent quantified nested Booleans:
Formulas $\iforall x. \, \, f$ and $\iexists x. \, f$ are represented as
$\iforall \, (\lamx{f})$ and $\iexists \, (\lamx{f})$. After clausification of
the input problem, no nested Booleans will be modified or renamed using fresh
predicate symbols.

The version of Zipperposition preceding our modifications distinguished between non\-equational
and equational literals. Following E \cite{scv-19-e23}, we
modified Zipperposition to represent all literals equationally: a nonequational
literal $f$ is stored as $f \eq \itrue$, whereas $\inot f$ is stored as $f
\noteq \itrue$. Equations of the form $f \eq \ifalse $ and $f \noteq \ifalse$ are
transformed into $f \noteq \itrue$ and $f \eq \itrue$, respectively.

@@

\looseness=-1
Rule \infname{PI} was already implemented by Simon Cruanes in Zipperposition,
before we started our modifications. Its implementation contains the following modes that generate
sets of possible terms $f$ for $P: \tau_1 \rightarrow \cdots \rightarrow \tau_n
\rightarrow o$: \emph{Full}, \emph{Pragmatic}, and $\mathit{Imit}_{\star}$. In  $\mathit{Imit}_{\star}$,
$\star$ is an element of a set of logical constants $S = \{ \iand, \ior,
\mathop{\ieq}\typeargs{\alpha}, \inot, \iforall, \iexists \}$. Mode \emph{Full}
generates imitations (for $P$) of all elements of $S$. Mode $\emph{Pragmatic}$
generates imitations of $\inot$, $\itrue$ and $\ifalse$; if there exist indices $i,j$
such that $i\not= j$ and  $\tau_i = \tau_j$, it generates $\lam{\tuplen{x}}{x_i
\ieq x_j}$; if there exist indices $i,j$ such that $i \not= j$, and $\tau_i =
\tau_j = o$, then it generates $\lam{\tuplen{x}}{x_i \iand x_j}$ and
$\lam{\tuplen{x}}{x_i \ior x_j}$; if for some $i$, $\tau_i = o$, then it
generates $\lam{\tuplen{x}}{x_i}$. Mode $\mathit{Imit}_\star$ generates
imitations of $\itrue$, $\ifalse$ and $\star$. In addition, $\mathit{Imit_{\iforall\iexists}}$ which generates imitations
of both $\iforall$ and $\iexists$.

@@

\looseness=-1
The axiom of choice is notoriously difficult to handle efficiently in higher-order provers. Andrews formulates
this axiom as $\iforall p.\, (\iexists
x.\, p \, x) \iimplies p \, (\varepsilon \, p)$, where $\varepsilon :
\forallty{\alpha}(\alpha \rightarrow o) \rightarrow \alpha$ denotes the \emph{choice
operator} \cite{pa-01-classical-ty-thy}. After clausification, this axiom becomes $P \, X \noteq \itrue \llor P
\, (\varepsilon \, P) \eq \itrue$. Since the term $P \, X$ matches any Boolean
term in the proof state, this axiom is very explosive. Therefore, Leo-III
\cite{sb-21-leo3} heuristically recognizes symbols that correspond to choice and deals with
them on the calculus level. Namely, whenever a clause $C = P \, X \noteq \itrue \llor P \, (\cst{f} \,
P) \eq \itrue$ is chosen for processing, $C$ is removed from the proof state and $\cst{f}$ is
added to set of choice functions $\mathit{CF}$ (which initially contains just
$\varepsilon$). Later, elements of $\mathit{CF}$ will be used to heuristically
instantiate the axiom of choice. We reused the method of recognizing choice
functions, but generalized the rule for creating the instance of
the axiom (assuming $\xi \in \mathit{CF}$ and $X$ and $z$ are fresh variables):
%
$$\namedinference{Choice}
               {C[\xi \, t]}
               { X \, (t \, Y) \noteq \itrue \llor X \, (t \, (\xi \, (\lam{z}{X \, (t \, z)})) ) \eq \itrue }$$
%
Let $D$ be the conclusion of \infname{Choice}. The fresh variable $X$ in $D$ acts as
arbitrary context around $t$, the chosen instantiation for $P$ from axiom of choice;
the variable $X$ can later be replaced by imitations of logical symbols to create more
complex instantiations of the choice axiom. To generate useful instances early, we create two instances: $\substcl{\{X
\mapsto \lam{z}{z}\}}{D}$ and $\substcl{\{X \mapsto \lam{z}{\inot\, z}\}}{D}$. Then, depending on
Zipperposition options, $D$ will either be deleted or kept. Note that $D$
will not subsume its instances, since the matching algorithm Zipperposition uses favors performance over completeness and is
thus too weak for this \cite[Sect.~6]{bbtvw-21-sup-lam}
\looseness=-1

Most provers natively support both functional and Boolean extensionality
reasoning: Bhayat et al. \cite{br-20-full-sup-w-combs} modify first-order
unification to return unification constraints consisting of pairs of terms of
functional type, whereas Steen relies on the unification rules of Leo-III's
calculus \cite[Section 4.3.3]{as-18-phd} to deal with extensionality. As a
pragmatic extension of the \lsup{} calculus, Bentkamp et al.
\cite{bbtvw-21-sup-lam} propose to alter the core generating inference rules of
superposition (Sect.~\ref{sec:pre:rules}) to support functional extensionality.
Instead of requiring that terms involved in the inference are unifiable, it is
required that they can be decomposed into \emph{disagreement pairs} such that at
least one of the disagreement pairs is of functional type. Disagreement pairs of
terms $s$ and $t$ of the same type are defined inductively using function
$\textsf{dp}$: $\textsf{dp}(s,t) = \emptyset$ if $s$ and $t$ are equal;
$\textsf{dp}(a \, \tuplen{s}, b \, \tuple{t}{m}) = \{(a \, \tuplen{s}, b \,
\tuple{t}{m})\}$ if $a$ and $b$ are different heads; $\textsf{dp}(\lam{x}{s},
\lam{y}{t}) = \{ (\lam{x}{s}, \lam{y}{t}) \}$; $\textsf{dp}(a \, \tuplen{s}, a
\, \tuplen{t}) = \bigcup_{i=1}^{n} \textsf{dp}(s_i, t_i)$. Then the
extensionality rules are stated as follows:
\pagebreak[2]
\begin{align*}
  & \namedinference{AbsSup}
  {s \eq t \llor C \qquad u[s'] \mathrel{\dot{\eq}} v \llor D}
  {\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \llor  u[t] \mathrel{\dot{\eq}} v \llor C \llor D}} \\
  &\namedinference{AbsER}{s \noteq s' \llor C}{\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \cdots \llor  C}} \\
  &\namedinference{AbsEF}{s \eq t \llor  s' \eq u \llor C}{\sigmacl{s_1 \noteq s'_1 \llor \cdots \llor s_n \noteq s'_n \llor t \noteq u \llor  s' \eq u \llor C}}
\end{align*}

\noindent\looseness=-1
In each of the rules, $\sigma$ is a most general unifier of the types of $s$ and
$s'$, and $\textsf{dp}(\sigmaterm{s} ,\sigmaterm{s'}) = \{ (s_1, s'_1), \ldots,
(s_n, s'_n)\}$. Rules \infname{AbsSup}, \infname{AbsER}, and \infname{AbsEF} are
extensional versions of superposition, equality resolution and equality
factoring (Sect.~\ref{sec:pre:rules}). All side conditions for extensional rules are the same as for
the standard rules, except that condition that $s$ and $s'$ are unifiable is
replaced by the condition that at least one $s_i$ is of functional type and that
$n>0$. By \infname{Abs} we denote the union of
these three rules. We extend \infname{Abs} to support Boolean extensionality by
requiring that at least one $s_i$ is of functional or type $o$, and adding the
condition ``$\textsf{dp}(f, g) = \{(f,g)\}$ if $f$ and $g$ are different
formulas'' to the definition of $\textsf{dp}$. If another condition of \textsf{dp}'s 
definition is also applicable, the newly added one is preferred. In the rest of the
text with \infname{Abs} we assume the version that supports Boolean extensionality.

\looseness=-1
Consider the clause $\cst{f} \, (\inot\cst{p} \ior \inot\cst{q}) \noteq \cst{f} \,
(\inot (\cst{p} \iand \cst{q}))$. This problem is unsatisfiable, as
the arguments of $\cst{f}$ on the different sides of the disequation are Boolean-extensionally
equal. Without the \infname{Abs} rules Zipperposition relies on
\infname{Cases}(\infname{Simp}) and \infname{IC} rules to derive the empty
clause. On the other hand, \infname{AbsER} generates~$C = \inot\cst{p} \ior \inot\cst{q} \noteq
\inot (\cst{p} \iand \cst{q})$. Then, $C$ gets clausified using
\infname{IC}, effectively reducing the problem to $\inot (\inot\cst{p} \ior
\inot\cst{q} \iequiv \inot (\cst{p} \iand \cst{q}))$, which is first-order.

@@

The winner of the higher-order theorem division of the 2019 edition of CASC \cite{gs-19-casc27}, Satallax \cite{cb-12-satallax}, has one
crucial advantage over Zipperposition: it is based on higher-order
tableaux, and as such it does not require formulas to be converted to clauses. The advantage of tableaux is that once it instantiates a variable with a
term, this instantiation naturally propagates through the whole formula. In
Zipperposition, which is based on \lsup{}, the original formula
is clausified and instantiating a variable in a clause $C$ does not automatically instantiate
it in all clauses that are results of clausification of the same formula as $C$.
To mitigate this issue, we have created extensions of equality resolution and equality factoring
that take Boolean extensionality into account:
%
\begin{align*}
  & \namedinference{BoolER}
  { s \eq s' \llor C}
  {\sigmacl{C}}
  &&
  \namedinference{BoolEF$+-$}
  {P\,\tuplen{s} \eq \itrue \llor s' \not\eq \itrue \llor C}
  {\sigmacl{s' \not\eq \itrue \llor C}} \\
  & \namedinference{BoolEF$-+$}
  {P\,\tuplen{s} \not\eq \itrue \llor s' \eq \itrue \llor C}
  {\sigmacl{s' \eq \itrue \llor C}} &&
  \namedinference{BoolEF$-$$-$}
  {P\,\tuplen{s} \not\eq \itrue \llor s' \not\eq \itrue \llor C}
  {\sigmacl{s' \not\eq \itrue \llor C}}
\end{align*}
%
All side conditions except for the ones concerning the unifiability of terms are
as in the original equality resolution and equality factoring rules. In rule
\infname{BoolER}, $\sigma$ is a unifier of $ s $ and $ \inot s' $. In the $+-$
and $-+$ versions of $\infname{BoolEF}$, $\sigma$ unifies $ P \, \tuplen{s}
$ and $\inot s'$, and in the remaining version it unifies $ P \, \tuplen{s} $ and
$s'$. Intuitively, these rules bring Boolean (dis)equations in the appropriate
form for application of the corresponding base rules. In particular,
for \infname{BoolER} we interpret the equation $s \eq s'$ as $s \not\eq \inot s'$ which
allows us to simulate \infname{ER} inference from Sect.~\ref{sec:pre:rules}. 
For the \infname{BoolEF} family of rules, we convert disequations into equations
using the trick of negating one side to simulate \infname{EF}.
% It suffices to consider 
% literals of the form $s \eq s'$ for \infname{BoolER} as Zipperposition rewrites
% $s \iequiv t \eq \itrue$ and $\inot(s \iequiv t) \not\eq \itrue$ to $s \eq t$ (and does analogous
% rewriting into $ s \not\eq t $).
The rule \infname{BoolER} only considers literals of the form $s \eq t$. 
Equivalences $s \iequiv t = \itrue$ and disequivalences $(s \iequiv t) \not\eq \itrue$ are automatically simplified to $s \eq t$ or $s \not\eq t$.
The example  \verb|SET557^1| in Sect.~\ref{sect:bool:examples} illustrates how 
the \infname{Bool} family of rules helps solve the problems that get obfuscated by clausification.

@@

\looseness=-1
In \infname{DC}$_\eq$ we require both $g$ and $h$ to be formulas and at least one of
them not to be $\itrue$. In \infname{DC}$_{\iforall}$, $Y$ is a fresh variable, and in
\infname{DC}$_{\iexists}$, $\cst{sk}$ is a fresh symbol and $\tuple{\alpha}{}$ and $\tuplen{Y}$ are all the type and term variables occurring
freely in $\iexists x.\, g$. The rules described above are as given by Ganzinger and Stuber (adapted to
our setting), with the omission of rules for negative literals (of the form $ f \not\eq
\itrue$), which are easy to derive and which can be found in their work \cite{gs-05-boolsup}.

\looseness=-1
Naive application of the \infname{DC} rules can result in exponential blowup in
problem size. To avoid this, we rename formulas $f$ that have repeated
occurrences by introducing predicates $\cst{p} \, \tuple{X}{n}$ replacing them, where
$\tuplen{X}$ are all free variables of $f$. We keep the count of all nonatomic
formulas occurring as either side of a literal. Before applying the \infname{DC}
rules on a clause $f \mathrel{\dot{\eq}} \itrue \llor C $, we check whether the
number of occurrences of $f$  exceeds the threshold $k$. If it does, based on the
polarity of the literal $f \mathrel{\dot{\eq}} \itrue$, we add the clause
$\cst{p} \, \tuplen{Y} \not\eq \itrue \llor f \eq \itrue$ (if the literal is
positive) or $\cst{p} \, \tuplen{Y} \eq \itrue \llor f \not\eq \itrue$ (if the
literal is negative), where $\tuplen{Y}$ are all free variables of $f$ and
$\cst{p}$ is a fresh symbol. Then, we replace the clause $f \mathrel{\dot\eq}
\itrue \llor C$ by $ \cst{p} \, \tuplen{Y} \mathrel{\dot\eq} \itrue \llor C$.

Before the number of occurrences of $f$ is checked, we first check (using a
fast, incomplete matching algorithm, used for simplification and subsumption) if there is a
formula $g$, for which definition was already introduced, such that
$\sigmaterm{g} = f$, for some substitution $\sigma$. This check can have three
outcomes. First, if the definition $\textsf{q} \, \overline{X}_n$ is already
introduced for $g$ with the same polarity as $f \, \dot{\eq} \,
\itrue$, then $f$ is replaced by $\sigmaterm{\cst{q} \, \tuplen{X}}$. Second, if
the definition was introduced, but with different polarity, we create the clause
defining $g$ with the missing polarity, and replace $f$ with $\sigmaterm{\cst{q}
\, \tuplen{X}}$. Last, if there is no renamed formula $g$ generalizing $f$,
we check if the number of occurrences of $f$ exceeds the threshold $k$.

@@

\looseness=-1
Rules \infname{Cases} and \infname{CasesSimp} deal with Boolean terms,
but we need to rely on extensionality reasoning to deal with $\lambda$-abstractions
whose body has type $o$. Using the observation that the formula $\iforall \tuplen{x}. \, f$ 
implies that $\lam{\tuplen{x}}{f}$ is functional-extensionally equal to $\lam{\tuplen{x}}{\itrue}$ (and similarly, if
$\iforall \tuplen{x}. \, \inot f$, then $\lam{\tuplen{x}}{f} \ieq \lam{\tuplen{x}}{\ifalse})$, we
designed the following rule (where $\tuplen{x}$ consists of all loosely bound variables of $f$ ):
%\pagebreak[2]
%
$$ \namedinference{Interpret$\lambda$}{C[\lam{\tuplen{x}}{f}]}
{(\iforall {\tuplen{x}}.\, f) \noteq \itrue \llor C[\lam{\tuplen{x}}{\itrue}] \qquad
(\iforall {\tuplen{x}}.\, \inot f) \noteq \itrue \llor C[\lam{\tuplen{x}}{\ifalse}]} $$
%

@@

The TPTP library \cite{gs-17-tptp} contains thousands of higher-order benchmarks, many of them hand-crafted
to point out subtle interferences of functional and Boolean properties of higher-order logic. In this section
we discuss some problems from the TPTP library that illustrate the advantages and disadvantages
of our approach.

@@

\looseness=-1
Combining these two rules with dynamic clausification is more useful: When rule
\infname{IC} is replaced by rule \infname{DC}, the success rate increases
to 1660 problems, compared to 1646 problems solved by b. We also discovered
that reasoning with choice is useful: When rule \infname{Choice} is enabled, the
success rate increases to 1653. We determined that including or excluding the
conclusion $D$ of \infname{Choice}, after it is simplified, makes no difference.
Counterintuitively, disabling \infname{BoolSimp} rule results in 1640 problems,
which is only 6 problems short of configuration b. Disabling \infname{Abs} and
\infname{Interpret$\lambda$} rules results in solving 25 and 31 problems less,
respectively. The raw data show that in total, using configurations from Figure
\ref{fig:cases} to Figure \ref{fig:bool-rules}, 1682 problems can be solved.

Last, we compare our approach to alternatives. Axiomatizing Booleans brings
Zipperposition down to a grinding halt: only 1106 problems can be solved using
this mode. On the other hand, preprocessing is fairly competitive: it solves
only 8 problems less than the b configuration.
